### 2014-11-04 experiments.tex
Add heat transfer. Add more results.

The measures of absolute run time performance for the VM were done
~\cite{cruz-ppdp14}.
To put our VM in perspective, we first compare it in terms of absolute

execution time with other competing systems using a single thread.

In Table~\ref{tbl:comp_nqueens}, we compare LM's version of the
classic N-Queens puzzle against 3 other versions: a straightforward
sequential program implemented in C using backtracking; a sequential
Python implementation~\cite{vanRossum:1995:PRM}; and a Prolog
implementation executed in YAP
Prolog~\cite{DBLP:journals/corr/abs-1102-3896}, an efficient
implementation of Prolog. Numbers less than 1 mean that LM is faster
and larger than 1 mean that LM is slower. We can observe that LM
easily beats Python, but is 5 to 10 times slower than YAP Prolog and
around 15 times slower than C. Note however that, as we will see next,
if we use at least 16 threads in LM, we can beat the sequential
implementation written in C.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{YAP Prolog} \\
\hline\hline
\textbf{10x10} & 16.92 & 0.62 & 5.42 \\
\textbf{11x11} & 21.59 & 0.64 & 6.47 \\
\textbf{12x12} & 10.32 & 0.73 & 7.61 \\
\textbf{13x13} & 14.35 & 0.88 & 10.38 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  N-Queens program}
\label{tbl:comp_nqueens}
\end{table}

In Table~\ref{tbl:comp_bp}, we compare LM's Belief Propagation (BP)
program, a machine learning algorithm to denoise images, against a
sequential C, Python and GraphLab~\cite{GraphLab2010} version of the
algorithm. GraphLab is a parallel C++ library used to solve
graph-based problems in machine learning. C and GraphLab perform about
the same since they are both compiled to machine code, although
the GraphLab version is highly optimized to run on multicore architectures.
Python runs very slowly since it is a
dynamic programming language and BP has many mathematical
computations. We should note, however, that LM's version uses some
external functions written in C++ in order to improve execution time,
therefore the comparison is not totally fair.

\begin{table}[ht]
\centering
{\begin{tabular}{c|c|c|c}
\textbf{Problem} & \multicolumn{3}{c}{\textbf{System}} \\
\textbf{Size} & \textbf{C} & \textbf{Python} & \textbf{GraphLab} \\
\hline\hline
\textbf{10}  & 1.00 & 0.03 & 1.00 \\
\textbf{50}  & 1.77 & 0.04 & 1.73 \\
\textbf{200} & 1.99 & 0.05 & 1.79 \\
\textbf{400} & 2.00 & 0.04 & 1.80 \\
\end{tabular}}
\caption{Comparing the absolute execution times (LM/System) for the
  Belief Propagation program}
\label{tbl:comp_bp}
\end{table}

We also compared a LM's version of the PageRank program against a
similar GraphLab version and LM showed to be around 4 to 6 times
slower. Our worse results were obtained for the all-pairs shortest
distance algorithm where a LM's version of the problem was around 50
times slower than a C sequential implementation of the Dijkstra
algorithm, but almost twice as fast when compared to the same
implementation in Python.

>>>>

In the paper~\cite{cruz-ppdp14}, the LM virtual machine was measured and
compared against implementations of similar programs but written in other
languages. For instance, the LBP program was found to be around 2 times slower
than GraphLab, while N Queens program was found to be around 10 to 15 slower
than a sequential C implementation. Some of those programs were compared against
a Python implementations and LM faired fairly well. In this section, we measure
the impact of the coordination mechanisms we have implemented for the new
virtual machine.

Fig.~\ref{results:comparison1}

\begin{figure}[h!]
   \begin{center}
      \subfloat[]{\includegraphics[width=9cm]{results/comparison1.png}}
   \end{center}
   \caption{Measuring the overhead of coordination mechanisms. The \textbf{Old}
   bars represent the performance of the original virtual machine, while
   \textbf{New} is the relative performance of the new version that includes
   coordination mechanisms. Most programs show little to no degradation in
   performance.}
   \label{results:comparison1}
\end{figure}

### 2014-11-04 heat-transfer.tex
Add heat transfer. Add more results.

Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
neighbor nodes. We added the first rule to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
node that changed significantly. With \texttt{set-priority} the node is
temporarily prioritized and selected to execute before others.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=*\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon * 2.0
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New),
         update(B), *underline[set-priority(B, Delta)]}.

@@

>>>>

Fig.~\ref{code:ht} shows the HT logical rules that send new heat values to
neighbor nodes. We added the first rule to increase the priority of the neighbor
nodes if the current node has a large $\delta$. The idea is to prioritize the
computation of heat values of nodes (using \texttt{update}) that have a neighbor
node that changed significantly. Multiple \texttt{add-priority} facts will
increase the priority of a node so that nodes with multiple large deltas will
have more priority. Note that the LM compiler will merge the three rules in
Fig.~\ref{code:ht} into a single one that handles the 3 branches.

\begin{figure}[h!]
\scriptsize\begin{Verbatim}[numbers=left,commandchars=\\\[\]]
new-heat(A, New, Old),
Delta = fabs(New - Old),
Delta > epsilon * 2.0
   -o {B, W | !edge(A, B, W) |
         new-neighbor-heat(B, A, New),
         update(B), \underline[add-priority(B, Delta)]}.

@@

Fig.~\ref{results:ht} presents the scalability results for the uncoordinated
and coordinated version. The dataset used is a square grid with an inner square
with high heat nodes. With 1 thread, there's a 33\% reduction in run time, while
for 16 threads, there's a 18\% reduction.

\begin{figure}[ht!]
   \begin{center}
      \subfloat[]{\includegraphics[width=5cm]{results/new-heat-transfer-80}}
   \end{center}
   \caption{Experimental results for the HT program. Speedup values for the
      coordinated and uncoordinated version are computed using the run time of
      the uncoordinated version using 1 thread.}
   \label{results:ht}
\end{figure}

### 2014-11-04 implementation.tex
More updates.
Add heat transfer. Add more results.

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.

@@

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.

Optimizations such as join optimizations (to allow rule filtering invalid
combinations) and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

@@

A node goes through different states during execution. The state machine in
Fig.~\ref{fig:node_states} represents the valid state transitions with the
following states:

@@

Threads pop active nodes from their queues and execute new candidate rules. 
Each thread has 2 queues, a doubly linked list known as the \emph{standard
queue} and a min/max heap known as the \emph{priority queue}. The regular queue
contains nodes without priorities and
implements the operations \texttt{push\_tail(node)} (push to the tail),
\texttt{pop\_head()} (remove node from the head),
\texttt{remove(node)} (remove an arbitrary node) and \texttt{remove\_half()}
(removes first half). The priority queue contains nodes
with priorities and is implemented as a binary heap array. It supports the
operations \texttt{push(node,priority)} (add new node to the heap),
           \texttt{pop\_min()} (remove the best node),
\texttt{remove(node)} (remove an arbitrary node), \texttt{pop\_half\_min()}
(remove the best half) and \texttt{move(node,new\_priority)} (update priority).
Operations such as \texttt{remove\_half()} are supported in order to support
node stealing, while operations \texttt{remove(node)} or
\texttt{move(node,new\_priority)} allows threads to change the priority of
nodes.

>>>>

The implementation of LM includes a compiler and a virtual machine that runs
byte-code. The virtual machine uses 32 registers for
operations and executes procedures to iterate over the database in order to match facts.
The implementation of the original machine is described in ~\cite{cruz-ppdp14}.
Here, we resume the most important details of that paper, while focusing on the
changes for efficiently implementing coordination.

@@

The compiler translates each rule to a procedure and a list of facts that need
to exist to satisfy the rule. This procedure is executed by the virtual
machine whenever we have enough facts to satisfy the rule. The procedure
loops over all possible combinations of the rule, retrieving facts from the
database, performing join operations and then consuming and deriving facts.
Optimizations such as join optimizations (to allow rule filtering invalid
combinations) and fact updates are implemented by the compiler. The latter
transforms fact derivations of the same predicate to a simple update operation,
avoiding a deallocation operation followed by an allocation of the new fact.

@@

In order to facilitate the implementation of coordination, we added a state
variable for each node. The state machine in
Fig.~\ref{fig:node_states} represents the valid state transitions of a node:

@@

Threads pop active nodes from their queues and execute new candidate rules. 
In order to allow node priorities, we use 2 queues per thread: a doubly linked
list known as the \emph{standard
queue} and a min/max heap known as the \emph{priority queue}. The regular queue
contains nodes without priorities and
implements the operations \texttt{push\_tail(node)} (push to the tail),
\texttt{pop\_head()} (remove node from the head),
\texttt{remove(node)} (remove an arbitrary node) and \texttt{remove\_half()}
(removes first half). The priority queue contains nodes
with priorities and is implemented as a binary heap array. It supports the
operations \texttt{push(node,priority)} (add new node to the heap),
           \texttt{pop\_min()} (remove the best node),
\texttt{remove(node)} (remove an arbitrary node), \texttt{pop\_half\_min()}
(remove the best half) and \texttt{move(node,new\_priority)} (update priority).
Operations such as \texttt{remove\_half()} are supported in order to support
node stealing, while operations \texttt{remove(node)} or
\texttt{move(node,new\_priority)} allows threads to change the priority of
nodes.
